Appendix

### 2025-04-09 manuscript_rmarkdown/manuscript_appendix.Rmd
Main changes to appendix are now complete.

---
title: "Appendix"
author: "Alex Zwanenburg, Steffen Löck"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: lualatex
    keep_tex: true
papersize: a4
bibliography: refs.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{amsthm}
- \usepackage{booktabs}
- \usepackage{multirow}
- \newtheorem{definition}{Definition}
- \DeclareMathOperator*{\argmax}{argmax}
- \DeclareMathOperator*{\argmin}{argmin}
- \DeclareMathOperator{\sgn}{sgn}
---

@@

# Appendix E: Simulations on clean data

Robust transformations are hypothesised to have a cost in efficiency for data
without outliers, i.e. clean data. Here we randomly draw data from a standard normal
distribution $\mathcal{N}(0,1)$. Subsequently, we perform an inverse transformation
$\left(\phi_{\text{YJ}}^{\lambda, 0, 1}\right)^{-1}$, with $\lambda \in \left(0, 2\right)$.

These transformed data

An alternative to location- and scale-invariant transformations is normalising feature distributions prior to conventional transformations.
We assessed two normalisation methods were used:
1. z-standardisation, i.e. $x^{\prime}_{i} = \left(x_i - \mu \right) / \sigma$, with $\mu$ and $\sigma$ the mean and standard deviation of feature $\mathbf{X}$; 
and 2. robust scaling, i.e. $x^{\prime}_{i} = \left(x_i - \text{median}\left(\mathbf{X}\right) \right) / \text{IQR}\left(\mathbf{X}\right)$, with $\text{IQR}$ representing the interquartile range of feature $\mathbf{X}$.

[Generated data: clean, n = 30, n = 100, n = 1000, lambda = 0.1, 1.0, 1.9]

```{=latex}
\begin{table}
\begin{center}
\caption{Residual errors for features from simulated clean data without outliers after Yeo-Johnson transformation to normality.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.}
\label{tab:clean-transformation-residuals-appendix}
\small{
\begin{tabular}{l | l r r r r r r r r r}

none                                  & & 4.0 & 3.4 & 4.6 & 40.1 & 8.9 & 20.8 & 184.6 & 10.1 & 191.8 \\
conventional                          & & 2.7 & 3.3 & 3.6 &  6.5 & 8.1 &  5.6 &  14.2 & 10.1 &  16.1 \\
conventional (z-score norm.)          & & 2.5 & 3.3 & 3.7 &  9.8 & 8.1 &  5.6 &  24.1 & 10.1 &  15.9 \\
conventional (robust scaling)         & & 2.6 & 3.3 & 3.7 &  7.4 & 8.2 &  5.6 &  15.5 & 10.1 &  14.0 \\
Raymaekers-Rousseeuw                  & & 3.8 & 3.3 & 3.6 &  6.5 & 8.5 &  5.6 &  14.4 & 11.0 &  17.6 \\
Raymaekers-Rousseeuw (z-score norm.)  & & 2.6 & 3.3 & 3.7 &  9.8 & 8.5 &  5.6 &  24.1 & 11.1 &  15.9 \\
Raymaekers-Rousseeuw (robust scaling) & & 2.6 & 3.3 & 3.7 &  7.4 & 8.6 &  5.6 &  15.4 & 11.1 &  15.1 \\
invariant                             & & 3.0 & 3.5 & 3.6 &  5.0 & 8.1 &  5.6 &  13.1 & 10.9 &  17.8 \\
robust invariant                      & & 3.0 & 3.3 & 3.6 &  5.2 & 8.3 &  5.9 &  19.9 & 13.6 &  32.9 \\
\bottomrule
\end{tabular}
}
\end{center}
\end{table}
```

```{=latex}
\begin{table}
\begin{center}
\caption{Transformation parameter $\lambda$ for features from simulated clean data without outliers after Yeo-Johnson transformation to normality.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.}
\label{tab:clean-transformation-lambda-appendix}
\small{
\begin{tabular}{l | l r r r r r r r r r}

conventional                          & & 0.6 & 1.2 & 1.5 &  0.0 & 0.9 & 1.6 &  0.1 & 1.0 & 1.9 \\
conventional (z-score norm.)          & & 0.5 & 1.2 & 1.5 & -0.2 & 0.9 & 1.7 & -0.2 & 1.0 & 2.2 \\
conventional (robust scaling)         & & 0.5 & 1.3 & 1.6 & -0.3 & 0.9 & 1.7 & -0.2 & 1.0 & 2.1 \\
Raymaekers-Rousseeuw                  & & 1.0 & 1.2 & 1.5 &  0.0 & 0.7 & 1.6 &  0.1 & 1.0 & 1.9 \\
Raymaekers-Rousseeuw (z-score norm.)  & & 0.7 & 1.2 & 1.5 & -0.2 & 0.7 & 1.7 & -0.2 & 1.0 & 2.2 \\
Raymaekers-Rousseeuw (robust scaling) & & 0.6 & 1.3 & 1.6 & -0.3 & 0.6 & 1.7 & -0.2 & 1.0 & 2.1 \\
invariant                             & & 0.4 & 1.4 & 1.7 &  0.1 & 0.9 & 1.8 &  0.1 & 1.1 & 3.0 \\
robust invariant                      & & 0.4 & 1.3 & 1.5 &  0.0 & 0.8 & 1.7 &  0.0 & 1.1 & 2.0 \\
\bottomrule
\end{tabular}
}
\end{center}
\end{table}
```

```{=latex}
\begin{table}
\begin{center}
\caption{
p-values of empirical central normality tests for features from simulated clean data without outliers after Yeo-Johnson transformation to normality.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.}
\label{tab:clean-transformation-p-value-appendix}
\small{
\begin{tabular}{p{3.5cm} | l r r r r r r r r r}

none                                  & & 0.986 & 0.614 & 0.323 & $<0.001$ & 0.063 & 0.016 & $<0.001$ & 0.997 & $<0.001$ \\
conventional                          & & 0.986 & 0.464 & 0.488 &   0.675  & 0.139 & 0.620 &   0.902  & 0.999 &   0.698 \\
conventional (z-score norm.)          & & 0.979 & 0.440 & 0.477 &   0.200  & 0.141 & 0.770 &   0.419  & 0.999 &   0.723 \\
conventional (robust scaling)         & & 0.982 & 0.439 & 0.484 &   0.487  & 0.130 & 0.683 &   0.883  & 0.999 &   0.817 \\
Raymaekers-Rousseeuw                  & & 0.989 & 0.464 & 0.488 &   0.675  & 0.446 & 0.620 &   0.876  & 0.992 &   0.821 \\
Raymaekers-Rousseeuw (z-score norm.)  & & 0.992 & 0.440 & 0.477 &   0.200  & 0.448 & 0.770 &   0.460  & 0.991 &   0.719 \\
Raymaekers-Rousseeuw (robust scaling) & & 0.993 & 0.439 & 0.484 &   0.487  & 0.422 & 0.683 &   0.887  & 0.991 &   0.848 \\
invariant                             & & 0.993 & 0.355 & 0.461 &   0.720  & 0.153 & 0.490 &   0.910  & 0.999 &   0.682 \\
robust invariant                      & & 0.994 & 0.407 & 0.484 &   0.642  & 0.291 & 0.489 &   0.848  & 0.996 &   0.628 \\
\bottomrule
\end{tabular}
}
\end{center}
\end{table}
```

@@

@@

@@

>>>>

---
title: "Appendix"
author: "Alex Zwanenburg, Steffen Löck"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: lualatex
    keep_tex: true
papersize: a4
bibliography: refs.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{amsthm}
- \usepackage{booktabs}
- \usepackage{multirow}
- \usepackage{placeins}
- \newtheorem{definition}{Definition}
- \DeclareMathOperator*{\argmax}{argmax}
- \DeclareMathOperator*{\argmin}{argmin}
- \DeclareMathOperator{\sgn}{sgn}
---

@@

\FloatBarrier

# Appendix E: Assessing transformations using simulated data

## Ranking Box-Cox transformations

Box-Cox transformations were assessed in the same manner as Yeo-Johnson transformations.
The results are shown in Table \ref{tab:comparison_methods_simulations_box_cox}.

@@

\FloatBarrier

## Examples using clean data

Robust transformations are hypothesised to have a cost in efficiency for data
without outliers, i.e. clean data. Here we draw nine sequences with elements randomly drawn from a standard normal
distribution $\mathcal{N}(0,1)$. Subsequently, we perform an inverse transformation
$\left(\phi^{\lambda, 0, 1}\right)^{-1}$, with $\lambda \in \left(0, 2\right)$.

The sequences drawn resulted from the permutations of the number of elements of each
sequence ($n \in \{30, 100, 500\}$) and transformation parameter for the
inverse transformation $\lambda \in \{0.1, 1.0, 1.9\}$. Each sequence then underwent
power transformation using Box-Cox and Yeo-Johnson transformations. For Box-Cox
transformations, each sequence was shifted prior to inverse transformation
so that every element after inverse transformation would be strictly positive.

Results for Box-Cox transformations are shown in Table
\ref{tab:clean-transformation-appendix-residuals-bc},
\ref{tab:clean-transformation-appendix-lambda-bc},
and \ref{tab:clean-transformation-appendix-p-value-bc}.
Additionally, results for Yeo-Johnson transformations are shown in Table
\ref{tab:clean-transformation-appendix-residuals-yj},
\ref{tab:clean-transformation-appendix-lambda-yj},
and \ref{tab:clean-transformation-appendix-p-value-yj}.
As may be observed, in these examples robust location and shift-invariant 
transformations have higher residual errors because of low weights being
assigned to tails of a distribution. This is also shown in 
Tables \ref{tab:comparison_methods_simulations_yeo_johnson}
and \ref{tab:comparison_methods_simulations_box_cox} for clean data without outliers.

\begin{table}
\begin{center}
\caption{Residual errors for features from simulated clean data without outliers after Yeo-Johnson transformation to normality.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.}
\label{tab:clean-transformation-appendix-residuals-yj}
\small{
\begin{tabular}{l | l r r r r r r r r r}

none                                  & & 4.0 & 3.4 & 4.6 & 40.1 &  8.9 & 20.8 & 184.6 & 10.1 & 191.8 \\
conventional                          & & 2.7 & 3.3 & 3.6 &  6.5 &  8.1 &  5.6 &  14.2 & 10.1 &  16.1 \\
conventional (z-score norm.)          & & 2.5 & 3.3 & 3.7 &  9.8 &  8.1 &  5.6 &  24.1 & 10.1 &  15.9 \\
conventional (robust scaling)         & & 2.6 & 3.3 & 3.7 &  7.4 &  8.2 &  5.6 &  15.5 & 10.1 &  14.0 \\
Raymaekers-Rousseeuw                  & & 3.8 & 3.3 & 3.6 &  6.5 &  8.5 &  5.6 &  14.4 & 11.0 &  17.6 \\
Raymaekers-Rousseeuw (z-score norm.)  & & 2.6 & 3.3 & 3.7 &  9.8 &  8.5 &  5.6 &  24.1 & 11.1 &  15.9 \\
Raymaekers-Rousseeuw (robust scaling) & & 2.6 & 3.3 & 3.7 &  7.4 &  8.6 &  5.6 &  15.4 & 11.1 &  15.1 \\
invariant                             & & 3.0 & 3.5 & 3.6 &  5.0 &  8.1 &  5.6 &  13.1 & 10.9 &  17.8 \\
robust invariant                      & & 3.0 & 3.4 & 3.6 &  5.5 & 10.1 &  6.2 &  18.7 & 13.2 &  32.8 \\

\bottomrule
\end{tabular}
}
\end{center}
\end{table}

\begin{table}
\begin{center}
\caption{Transformation parameter $\lambda$ for features from simulated clean data without outliers after Yeo-Johnson transformation to normality.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.}
\label{tab:clean-transformation-appendix-lambda-yj}
\small{
\begin{tabular}{l | l r r r r r r r r r}

conventional                          & & 0.6 & 1.2 & 1.5 &  0.0 & 0.9 & 1.6 &  0.1 & 1.0 & 1.9 \\
conventional (z-score norm.)          & & 0.5 & 1.2 & 1.5 & -0.2 & 0.9 & 1.7 & -0.2 & 1.0 & 2.2 \\
conventional (robust scaling)         & & 0.5 & 1.3 & 1.6 & -0.3 & 0.9 & 1.7 & -0.2 & 1.0 & 2.1 \\
Raymaekers-Rousseeuw                  & & 1.0 & 1.2 & 1.5 &  0.0 & 0.7 & 1.6 &  0.1 & 1.0 & 1.9 \\
Raymaekers-Rousseeuw (z-score norm.)  & & 0.7 & 1.2 & 1.5 & -0.2 & 0.7 & 1.7 & -0.2 & 1.0 & 2.2 \\
Raymaekers-Rousseeuw (robust scaling) & & 0.6 & 1.3 & 1.6 & -0.3 & 0.6 & 1.7 & -0.2 & 1.0 & 2.1 \\
invariant                             & & 0.4 & 1.4 & 1.7 &  0.1 & 0.9 & 1.8 &  0.1 & 1.1 & 3.0 \\
robust invariant                      & & 0.5 & 1.1 & 1.5 & -0.1 & 0.7 & 1.7 &  0.1 & 1.1 & 1.9 \\
\bottomrule
\end{tabular}
}
\end{center}
\end{table}

\begin{table}
\begin{center}
\caption{
p-values of empirical central normality tests for features from simulated clean data without outliers after Yeo-Johnson transformation to normality.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.}
\label{tab:clean-transformation-appendix-p-value-yj}
\small{
\begin{tabular}{p{3.5cm} | l r r r r r r r r r}

none                                  & & 0.981 & 0.816 & 0.576 & $<0.001$ & 0.099 & 0.011 & $<0.001$ & 0.993 & $<0.001$ \\
conventional                          & & 0.979 & 0.614 & 0.617 &   0.722  & 0.221 & 0.782 &   0.946  & 0.997 &   0.637 \\
conventional (z-score norm.)          & & 0.970 & 0.594 & 0.645 &   0.251  & 0.224 & 0.895 &   0.529  & 0.998 &   0.828 \\
conventional (robust scaling)         & & 0.974 & 0.593 & 0.640 &   0.538  & 0.207 & 0.837 &   0.933  & 0.998 &   0.810 \\
Raymaekers-Rousseeuw                  & & 0.983 & 0.614 & 0.617 &   0.722  & 0.622 & 0.782 &   0.927  & 0.985 &   0.831 \\
Raymaekers-Rousseeuw (z-score norm.)  & & 0.985 & 0.594 & 0.645 &   0.251  & 0.623 & 0.895 &   0.564  & 0.984 &   0.807 \\
Raymaekers-Rousseeuw (robust scaling) & & 0.987 & 0.593 & 0.640 &   0.538  & 0.599 & 0.837 &   0.935  & 0.984 &   0.875 \\
invariant                             & & 0.987 & 0.487 & 0.573 &   0.778  & 0.241 & 0.683 &   0.963  & 0.997 &   0.865 \\
robust invariant                      & & 0.992 & 0.753 & 0.667 &   0.909  & 0.739 & 0.628 &   0.968  & 0.994 &   0.680 \\

\bottomrule
\end{tabular}
}
\end{center}
\end{table}

@@

\begin{table}
\begin{center}
\caption{Residual errors for features from simulated clean data without outliers after Box-Cox transformation to normality.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.}
\label{tab:clean-transformation-appendix-residuals-bc}
\small{
\begin{tabular}{l | l r r r r r r r r r}

none                                  & & 5.3 & 3.4 & 3.8 & 45.7 &  8.9 & 6.0 & 221.8 & 10.1 & 29.7 \\
conventional                          & & 2.4 & 3.2 & 3.5 &  5.9 &  8.4 & 5.5 &  14.3 &  9.8 & 16.1 \\
conventional (z-score norm.)          & & 2.4 & 3.2 & 3.7 &  5.4 &  8.7 & 5.5 &  14.0 & 10.4 & 16.1 \\
conventional (robust scaling)         & & 2.4 & 3.2 & 3.7 &  5.4 &  8.7 & 5.5 &  14.0 & 10.4 & 16.1 \\
Raymaekers-Rousseeuw                  & & 2.4 & 3.2 & 3.5 &  5.9 & 13.3 & 5.7 &  16.1 & 13.6 & 16.9 \\
Raymaekers-Rousseeuw (z-score norm.)  & & 2.4 & 3.2 & 4.9 &  6.3 &  8.9 & 6.2 &  19.1 & 19.0 & 19.9 \\
Raymaekers-Rousseeuw (robust scaling) & & 2.4 & 3.2 & 4.9 &  6.3 &  8.9 & 6.2 &  18.4 & 19.0 & 19.9 \\
invariant                             & & 2.4 & 3.2 & 3.7 &  5.4 &  8.7 & 5.5 &  13.7 &  9.9 & 16.0 \\
robust invariant                      & & 2.4 & 4.3 & 4.7 &  8.9 & 20.4 & 6.7 &  14.1 & 15.8 & 37.6 \\

\bottomrule
\end{tabular}
}
\end{center}
\end{table}

\begin{table}
\begin{center}
\caption{Transformation parameter $\lambda$ for features from simulated clean data without outliers after Box-Cox transformation to normality.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.}
\label{tab:clean-transformation-appendix-lambda-bc}
\small{
\begin{tabular}{l | l r r r r r r r r r}

conventional                          & & 0.4 & 1.3 & 0.5 & 0.0 &  0.9 & 0.7 & 0.1 & 1.1 & 2.0 \\
conventional (z-score norm.)          & & 0.4 & 1.1 & 0.8 & 0.2 &  0.9 & 0.8 & 0.1 & 1.0 & 1.4 \\
conventional (robust scaling)         & & 0.4 & 1.1 & 0.8 & 0.2 &  0.9 & 0.8 & 0.1 & 1.0 & 1.4 \\
Raymaekers-Rousseeuw                  & & 0.4 & 1.3 & 0.5 & 0.0 & -0.1 & 0.5 & 0.1 & 0.9 & 1.8 \\
Raymaekers-Rousseeuw (z-score norm.)  & & 0.4 & 1.1 & 0.4 & 0.1 &  0.8 & 0.7 & 0.1 & 0.9 & 1.2 \\
Raymaekers-Rousseeuw (robust scaling) & & 0.4 & 1.1 & 0.4 & 0.1 &  0.8 & 0.7 & 0.1 & 0.9 & 1.2 \\
invariant                             & & 0.4 & 1.6 & 0.8 & 0.2 &  0.9 & 0.8 & 0.1 & 1.1 & 1.9 \\
robust invariant                      & & 0.5 & 0.8 & 0.4 & 0.0 &  0.2 & 0.7 & 0.1 & 0.9 & 0.9 \\

\bottomrule
\end{tabular}
}
\end{center}
\end{table}

\begin{table}
\begin{center}
\caption{
p-values of empirical central normality tests for features from simulated clean data without outliers after Box-Cox transformation to normality.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.}
\label{tab:clean-transformation-appendix-p-value-bc}
\small{
\begin{tabular}{p{3.5cm} | l r r r r r r r r r}

none                                  & & 0.940 & 0.816 & 0.498 & $<0.001$ & 0.098 & 0.714 & $<0.001$ & 0.993 & 0.928 \\
conventional                          & & 0.948 & 0.695 & 0.603 &   0.798  & 0.163 & 0.791 &   0.931  & 0.997 & 0.658 \\
conventional (z-score norm.)          & & 0.947 & 0.777 & 0.624 &   0.611  & 0.182 & 0.788 &   0.969  & 0.991 & 0.822 \\
conventional (robust scaling)         & & 0.947 & 0.777 & 0.624 &   0.611  & 0.182 & 0.788 &   0.969  & 0.991 & 0.822 \\
Raymaekers-Rousseeuw                  & & 0.948 & 0.695 & 0.603 &   0.798  & 0.768 & 0.806 &   0.816  & 0.962 & 0.782 \\
Raymaekers-Rousseeuw (z-score norm.)  & & 0.947 & 0.777 & 0.753 &   0.913  & 0.255 & 0.784 &   0.775  & 0.866 & 0.921 \\
Raymaekers-Rousseeuw (robust scaling) & & 0.947 & 0.777 & 0.753 &   0.913  & 0.255 & 0.784 &   0.807  & 0.866 & 0.921 \\
invariant                             & & 0.947 & 0.661 & 0.624 &   0.611  & 0.182 & 0.789 &   0.956  & 0.997 & 0.664 \\
robust invariant                      & & 0.969 & 0.824 & 0.750 &   0.972  & 0.779 & 0.775 &   0.956  & 0.934 & 0.840 \\

\bottomrule
\end{tabular}
}
\end{center}
\end{table}

\FloatBarrier

@@

\FloatBarrier
